Abstract Data Type Inference

### 2015-03-11 inference.tex
Better Section 2.

We fix an arbitrary infinite set $\mathbb{O}$ of operation identifiers, and
given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and values. A
\emph{call action} binds an operation identifier $o \in \mathbb{O}$ with a
method name $p \in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a
\emph{return action} binds an operation identifier $o$ with a return value $v$.
An \emph{execution} $e$ is a sequence of call and return actions where
\begin{itemize}

@@

  \item permuting call actions backward in time, and

  \item permuting return actions forward in time.

@@

In this work, histories are abstractions of executions which retain method
names yet ignore argument and return values, and retain the relative order of
operations yet ignore the exact sequence of call and return actions. Formally,
a \emph{history} $h = \tup{O,<,c,f,m}$ is a tuple where
\begin{itemize}

@@

\end{itemize}
Non-completed operations are \emph{pending}, and are maximal in the
happens-before order. The \emph{width} of $h$ is the maximal size of a
subset of $O$ for which each pair of operations is unordered. The \emph{width}
of a history set is the maximal width of its elements. Width-$1$ histories are
\emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a
\emph{match}. We assume targets target themselves: $o \in \set{ o' \in O :
m(o') = o}$.

Our abstraction from executions to histories relies correlating the operations
associated with the same values via matching. A \emph{matching scheme} $M$
associates to each execution $e$ with operations $O$ a partial matching
function $M(e): O \rightharpoonup O$.

@@

The history $H(M,e)$ of an execution $e$ under matching scheme $M$ is the tuple
$\tup{O,<,c,f,M(e)}$ where
\begin{itemize}

@@

\end{itemize}
We denote the set $\set{H(M,e) : e \in E}$ of histories of an execution set $E$
by $H(M,E)$.

A matching scheme $M$ is \emph{faithful} to a set $E$ of executions when $e \in
E$ if{f} $e' \in E$ for any two executions $e$ and $e'$ such that $H(M,e) =
H(M,e')$. A set of executions $E$ is \emph{data independent}\footnote{TODO NOTE
THAT THIS IS NOT THE STANDARD DEFINITION…} when there exists a faithful
matching scheme. By definition, abstracting executions to histories comes at no
loss of precision up to set inclusion for data-independent execution sets.

@@

  $H(M,e) \in H(M,E)$ if and only if $e \in E$, for any faithful matching
  scheme $M$.

@@

In Section~\ref{sec:nature} we demonstrate faithful matching schemes for the
executions of naturally-occurring concurrent objects, and in
Section~\ref{sec:matching} we demonstrate how to infer faithful matching
schemes. Otherwise, for the remainder of this work, we assume each set of
executions comes equipped with a faithful matching scheme $M$, and abbreviate
$H(M,E)$ by $H(E)$.

@@

An \emph{abstract data type (ADT)} $A$ is the set $H(\mathcal{I})$ of histories
of an implementation $\mathcal{I}$. A \emph{kernel} of an ADT $A$ is a minimal
set $H$ of histories such that
\begin{align*}
  A = \set{ h' : \exists h \in H.\ h \to^\ast h' }
\end{align*}
where $\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$. In
Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
unique kernels. For the remainder of this work, we assume that ADTs have unique
kernels, and denote \emph{the kernel} of $A$ by $\ker A$. We say that $A$ is
\emph{atomic} when the width of $\ker A$ is $1$.

We define \emph{observational refinement} between two implementations, or
between an implementation and an ADT, as history-set inclusion. We say
$\mathcal{I}_1$ \emph{refines} $\mathcal{I}_2$ if{f} $H(\mathcal{I}_1)
\subseteq H(\mathcal{I}_2)$, and $\mathcal{I}$ \emph{refines} $A$ if{f}
$H(\mathcal{I}) \subseteq A$.

In the following we fix a symbolic representation of ADTs on which to base our
symbolic ADT inference problem. A \emph{history formula} $F$ is a first-order
logic formula with
\begin{itemize}

@@

The \emph{complement} of an ADT $A$ with width $k \in \mathbb{N}$ is the set of
histories of width at most $k$ which are excluded from $A$. A history formula
$F$ \emph{represents} an ADT $A$ when:
\begin{itemize}

@@

  \item $h \not\models F$ for all $h \in B$,

\end{itemize}
where $B$ is the complement of $\ker A$. The \emph{symbolic ADT inference
problem} is to compute a history formula $F$ representing the ADT of a given
implementation $\mathcal{I}$.

>>>>

We fix an arbitrary infinite set $\mathbb{O}$ of operation identifiers, and
given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and values. A
\emph{call action} binds an operation $o \in \mathbb{O}$ with a method name $p
\in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a \emph{return
action} binds an operation $o \in \mathbb{O}$ with a return value $v \in
\mathbb{V}$. An \emph{execution} $e$ is a sequence of call and return actions
where
\begin{itemize}

@@

  \item permuting call actions backward, and

  \item permuting return actions forward.

@@

In this work, histories are abstractions of executions which retain method
names yet lose the exact argument and return values, and retain the relative
order of operations yet lose the exact sequence of call and return actions.
Formally, a \emph{history} $h = \tup{O,<,c,f,m}$ is a tuple where
\begin{itemize}

@@

\end{itemize}
Non-completed operations are \emph{pending}, and are maximal in the
happens-before order. The \emph{width} of a history is the maximum number of
its operations which are mutually unordered, and the \emph{width} of a history
set is the maximum width of its elements. Histories with width $1$ are
\emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a
\emph{match}. We assume targets target themselves: $m(o) = o$ for $o \in
\rng(m)$. The \emph{weight} of $h$ is the number $|\rng(m)|$ of matches.
Since the operation identifiers of a history have no intrinsic meaning, we
consider equality between histories up to renaming of operation identifiers.

  TODO HISTORY EXAMPLE

Our abstraction of executions relies on correlating the operations associated
with the same values in an execution via the partial matching function of a
history. We construct partial matching functions systematically. A
\emph{matching scheme} $M$ associates to each execution $e$ with operations $O$
a partial matching function $M(e): O \rightharpoonup O$.

@@

The history $H(e,M)$ of an execution $e$ under matching scheme $M$ is the tuple
$\tup{O,<,c,f,M(e)}$ where
\begin{itemize}

@@

\end{itemize}
We denote the set $\set{H(e,M) : e \in E}$ of histories of an execution set $E$
by $H(E,M)$.

  TODO HISTORY OF AN EXECUTION

A matching scheme $M$ is \emph{faithful} to a set $E$ of executions when $e \in
E$ if{f} $e' \in E$ for any two executions $e$ and $e'$ such that $H(e,M) =
H(e',M)$. A set of executions $E$ is \emph{data independent}\footnote{TODO NOTE
THAT THIS IS NOT THE STANDARD DEFINITION…} when there exists a faithful
matching scheme. By definition, our abstraction of executions incurs no loss of
precision for data-independent execution sets.

@@

  $H(e,M) \in H(E,M)$ if and only if $e \in E$, for any faithful matching
  scheme $M$.

@@

In Section~\ref{sec:nature} we demonstrate faithful matching schemes for the
executions of naturally-occurring ADTs, and in Section~\ref{sec:matching} we
demonstrate how to infer faithful matching schemes. Otherwise, for the
remainder of this work, we assume each set of executions comes equipped with a
faithful matching scheme $M$, and abbreviate $H(E,M)$ by $H(E)$.

@@

An \emph{abstract data type (ADT)} $A$ is the set $H(\mathcal{I})$ of histories
of an implementation $\mathcal{I}$. A \emph{kernel} of an ADT $A$ is a minimal
set $H$ of histories generating $A$, i.e.,~such that
\begin{align*}
  A = \set{ h' : \exists h \in H.\ h \to^\ast h' }
\end{align*}
where $\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$. In
Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
unique kernels. For the remainder of this work, we assume that ADTs have unique
kernels, and denote \emph{the kernel} of $A$ by $\ker A$. We say that $A$ is
\emph{atomic} when the width of $\ker A$ is $1$.

We define \emph{observational refinement} between implementations and ADTs as
history-set inclusion. Although this refinement is typically defined with
respect to the admissibility of program executions, recent work demonstrates
that these definitions are equivalent~\cite{conf/popl/BouajjaniEEH15}.

\begin{definition}

  We say an implementation $\mathcal{I}$ \emph{refines} an ADT $A$ when
  $H(\mathcal{I}) \subseteq A$. We say an implementation $\mathcal{I}_1$
  \emph{refines} another implementation $\mathcal{I}_2$ when $\mathcal{I}_1$
  refines $H(\mathcal{I}_2)$.

\end{definition}

In order to frame the symbolic ADT inference problem, we fix a natural symbolic
representation of ADTs. A \emph{history formula} $F$ is a first-order logic
formula with
\begin{itemize}

@@

The \emph{complement} of a history set $H$ of width $k \in \mathbb{N} \cup
\set{\omega}$ is the set of histories of width at most $k$ which are excluded
from $H$. Let $A$ be an ADT and $B$ the complement of $\ker A$. We say that
a history formula $F$ \emph{represents} $A$ when
\begin{itemize}

@@

  \item $h \not\models F$ for all $h \in B$.

\end{itemize}
The inference problem is thus to compute a formula representing an ADT.

\begin{definition}

  The \emph{symbolic ADT inference problem} is to compute a history formula $F$
  representing the ADT $H(\mathcal{I})$ of a given implementation $\mathcal{I}$.

\end{definition}
